New cardinality estimation algorithms for HyperLogLog sketches

### 2016-12-28 paper/paper.tex
some minor improvements

\begin{algorithm}
\caption{Compression of a $(\symPrecision, \symRegRange)$-HyperLogLog sketch with register values $\boldsymbol{\symRegValVariate}=(\symRegValVariate_1,\ldots,\symRegValVariate_\symNumReg)$ into a $(\symPrecision', \symRegRange')$-HyperLogLog sketch with $\symPrecision'\leq\symPrecision$ and $\symPrecision'+\symRegRange' \leq\symPrecision + \symRegRange$.}
\label{alg:compress}
\begin{algorithmic}
\Function {Compress}{\symDataItem, $\boldsymbol{\symRegValVariate}$}
\State allocate registers for new HyperLogLog sketch $\boldsymbol{\symRegValVariate}' = (\symRegValVariate_1,\ldots,\symRegValVariate_{2^{\symPrecision'}})$
\For{$\symIndexI\gets 1, 2^{\symPrecision'}$}
\State $\symB \gets (\symIndexI-1)\cdot 2^{\symPrecision-\symPrecision'}$
\State $\symIndexJ\gets 1$
\While{$\symIndexJ \leq 2^{\symPrecision-\symPrecision'}
\wedge
\symRegValVariate_{\symB+\symIndexJ}=0
$}
\State $\symIndexJ\gets\symIndexJ+1$
\EndWhile
\If{$\symIndexJ = 1$}
\State $\symRegValVariate'_\symIndexI \gets \min(\symRegValVariate_{\symB+\symIndexJ} + \symPrecision-\symPrecision',\symRegRange'+1)$
\ElsIf{$\symIndexJ\leq2^{\symPrecision-\symPrecision'}$}
\State $\langle \symBitRepA_1, \ldots, \symBitRepA_{\symPrecision-\symPrecision'}\rangle_2 \gets \symIndexJ-1$
\State $\symRegValVariate'_\symIndexI \gets 
\min(\{\symS\mid \symS \in [1, \symPrecision-\symPrecision']  \wedge  \symBitRepA_\symS = 1\} \cup \{\symRegRange'+1\})$
\Else
\State $\symRegValVariate'_\symIndexI \gets 0$
\EndIf
\EndFor
\State \Return $\boldsymbol{\symRegValVariate}'$
\EndFunction
\end{algorithmic}
\end{algorithm}

@@

The original estimation algorithm as presented in \cite{Flajolet2007} including corrections for small and large cardinalities is summarized by \cref{alg:estimate_original}. 
\begin{algorithm}
\caption{Original cardinality estimation algorithm for HyperLogLog sketches that use 32-bit hash values ($\symPrecision+\symRegRange = 32$) for insertion of data items \cite{Flajolet2007}.}
\label{alg:estimate_original}
\begin{algorithmic}
\Function {EstimateCardinality}{$\boldsymbol{\symRegValVariate}$}
\State $\symNumReg \gets \dim \boldsymbol{\symRegValVariate}$
\State $\symCardinalityRawEstimate = \symAlpha_\symNumReg \symNumReg^2\left(\sum_{\symIndexI=1}^\symNumReg 2^{-\symRegValVariate_\symIndexI}\right)^{\!-1}$
\comm{.4}{raw estimate \eqref{equ:raw_estimator}}
\If{$\symCardinalityRawEstimate\leq \frac{5}{2}\symNumReg$}
\State $\symCountVariate_0 = \left|\{\symIndexI \vert\symRegValVariate_\symIndexI=0\}\right|$
\If{$\symCountVariate_0\neq0$}
\State \Return $\symNumReg \log(\symNumReg / \symCountVariate_0)$
\comm{.4}{small range correction \eqref{equ:linear_counting_estimator}}
\Else
\State \Return $\symCardinalityRawEstimate$
\EndIf
\ElsIf{$\symCardinalityRawEstimate\leq \frac{1}{30}2^{32}$}
\State \Return $\symCardinalityRawEstimate$
\Else
\State\Return $-2^{32}\log(1-\symCardinalityRawEstimate/2^{32})$
\comm{.4}{large range correction \eqref{equ:large_range_estimate}}
\EndIf
\EndFunction
\end{algorithmic}
\end{algorithm}
The relative estimation error for the original method is shown in \cref{fig:original_estimate}. Unfortunatley, as can be clearly seen, the ranges where the estimation error is small for $\symCardinalityRawEstimate$ and $\symCardinalityEstimate_\text{small}$ do not overlap. Therefore, the estimation error is much larger near the transition region. To reduce the estimation error for cardinalities close to this region, it was proposed to correct $\symCardinalityRawEstimate$ for bias. Empirically collected bias correction data is either stored as set of interpolation points \cite{Heule2013}, as lookup table \cite{Rhodes2015}, or as best-fitting polynomial \cite{Sanfilippo2014}. However, all these empirical approaches treat the symptom and not the cause.

@@

We can cross-check the new estimator for the linear counting case with $\symRegRange=0$. Using the identity $\symSmallCorrectionFunc(\symX) + \symLargeCorrectionFunc(\symX) = \symAlpha_\infty\symPowerSeriesFunc(\log_2(\log(1/\symX)))/\log(1/\symX)$, we get
\begin{equation}
\symPoissonRateEstimate 
= 
\frac{\symAlpha_\infty\symNumReg}{
\symSmallCorrectionFunc(\symCountVariate_0/\symNumReg)
+
\symLargeCorrectionFunc(\symCountVariate_0/\symNumReg)
}
=
\frac{\symNumReg\log\!\left(\symNumReg/\symCountVariate_0\right)}{\symPowerSeriesFunc\!\left(\log_2\!\left(\log\!\left(\symNumReg/\symCountVariate_0\right)\right)\right)}
\end{equation}
which is as expected almost identical to the linear counting estimator \eqref{equ:linear_counting_estimator}, because $\symPowerSeriesFunc(\symX)\approx 1$ (see \cref{sec:derivation_raw_estimator}).
\subsection{Improved raw estimation algorithm}
The improved raw estimator \eqref{equ:correctedestimator} leads to a new cardinality estimation algorithm for HyperLogLog sketches. \cref{alg:corrected_raw_estimation} demonstrates the numerical evaluation of the estimator. The values of functions $\symSmallCorrectionFunc$ and $\symLargeCorrectionFunc$ can be either calculated on-demand or pre-calculated. The possible value range of $\symCountVariate_0$ and $\symCountVariate_{\symRegRange+1}$ is $[0, \symNumReg]$. Therefore, if performance matters, the function values can be precalculated for all possible arguments and kept in lookup tables of size $\symNumReg+1$.
The new estimation algorithm is very elegant, because it does neither contain magic numbers nor special cases as the original algorithm.

\begin{algorithm}
\caption{Cardinality estimation algorithm based on the improved raw estimator.}
\label{alg:corrected_raw_estimation}
\begin{algorithmic}
\Function {EstimateCardinality}{$\boldsymbol{\symCountVariate}$}
\State $\symNumReg \gets \left\Vert\boldsymbol{\symCountVariate}\right\Vert_1$
\State $\symZ\gets\symNumReg\cdot\symLargeCorrectionFunc(1 - \symCountVariate_{\symRegRange+1}/\symNumReg)$
\comm{.5}{alternatively, take $\symNumReg\cdot\symLargeCorrectionFunc(1 - \symCountVariate_{\symRegRange+1}/\symNumReg)$ from precalculated lookup table}
\For{$\symRegVal\gets \symRegRange, 1$}
\State $\symZ\gets0.5\cdot\left(\symZ + \symCountVariate_\symRegVal\right)$
\EndFor
\State $\symZ\gets\symZ+\symNumReg\cdot\symSmallCorrectionFunc(\symCountVariate_0/\symNumReg)$
\comm{.5}{alternatively, take $\symNumReg\cdot\symSmallCorrectionFunc(\symCountVariate_0/\symNumReg)$ from precalculated lookup table}
\State \Return$\symAlpha_\infty\symNumReg^2 / \symZ$
\comm{.5}{$\symAlpha_\infty := 1/(2\log(2))$}
\EndFunction
\\
\Function {$\symSmallCorrectionFunc$}{$\symX$}
\comm{.5}{$\symX\in[0,1]$}
\If{$\symX=1$}
\State\Return$\infty$
\EndIf
\State $\symY\gets1$
\State $\symZ\gets\symX$
\Repeat
\State $\symX\gets\symX\cdot\symX$
\State $\symZ'\gets\symZ$
\State $\symZ\gets\symZ + \symX\cdot\symY$
\State $\symY\gets2\cdot\symY$
\Until{$\symZ=\symZ'$}
\State\Return$\symZ$
\EndFunction
\\
\Function {$\symLargeCorrectionFunc$}{$\symX$}
\comm{0.5}{$\symX\in[0,1]$}
\State $\symY\gets1$
\State $\symZ\gets0$
\Repeat
\State $\symX\gets\sqrt{\symX}$
\State $\symZ'\gets\symZ$
\State $\symY\gets0.5\cdot\symY$
\State $\symZ\gets\symZ + \left(1-\symX\right)\cdot\symX\cdot\symY$
\Until{$\symZ=\symZ'$}
\State\Return$\symZ$
\EndFunction
\end{algorithmic}
\end{algorithm}

@@

\begin{algorithm}
\caption{Maximum likelihood cardinality estimation.}
\ContinuedFloat
\label{alg:estimate_ml}
\begin{algorithmic}
\Function {EstimateCardinality}{$\boldsymbol{\symCountVariate}$}
\State $\symRegRange\gets\dim(\boldsymbol{\symCountVariate})-2$
\State $\symRegValVariate_\text{min} \gets \min\lbrace \symRegVal\vert\symCountVariate_\symRegVal>0\rbrace$
\If{$\symRegValVariate_\text{min} > \symRegRange$}
\State\Return $\infty$
\EndIf
\State $\symRegValVariate'_\text{min} \gets \max(\symRegValVariate_\text{min}, 1)$
\State $\symRegValVariate_\text{max} \gets \max\lbrace \symRegVal\vert\symCountVariate_\symRegVal>0\rbrace$
\State $\symRegValVariate'_\text{max} \gets \min(\symRegValVariate_\text{max}, \symRegRange)$
\State $\symZ \gets 0$
\State $\symNumReg'\gets \symCountVariate_{\symRegRange+1}$
\State $\symY \gets 2^{-\symRegValVariate'_\text{max}}$
\For{$\symRegVal \gets \symRegValVariate'_\text{max},\symRegValVariate'_\text{min}$}
\State $\symZ \gets \symZ + \symCountVariate_\symRegVal\cdot\symY$
\comm{0.35}{here $\symY = 2^{-\symRegVal}$}
\State $\symY \gets 2\symY$
\State $\symNumReg'\gets\symNumReg' + \symCountVariate_\symRegVal$
\EndFor
\State
\comm{0.35}{here $\symZ=\sum_{\symRegVal=1}^{\symRegRange}\frac{\symCountVariate_\symRegVal}{2^\symRegVal}$}
\State $\symNumReg \gets \symNumReg' + \symCountVariate_0$ 
\State $\symCount \gets \symCountVariate_{\symRegRange+1}$
\If{$\symRegRange\geq 1$}
\State $\symCount \gets \symCount + \symCountVariate_{\symRegValVariate'_\text{max}}$
\EndIf
\State $\symFuncPrime_\text{prev}\gets 0$
\State $\symA\gets \symZ + \symCountVariate_0$
\comm{0.35}{$\symA = \sum_{\symRegVal=0}^{\symRegRange}\frac{\symCountVariate_\symRegVal}{2^\symRegVal}$}
\State $\symB\gets \symZ + 
\symCountVariate_{\symRegRange+1}\cdot 2^{-\symRegRange}$
\comm{0.35}{$\symB = \sum_{\symRegVal=1}^{\symRegRange}\frac{\symCountVariate_\symRegVal}{2^\symRegVal} + \frac{\symCountVariate_{\symRegRange+1}}{2^\symRegRange}$}
\If{$\symB \leq 1.5\cdot\symA$}
\State $\symX \gets \symNumReg'/(0.5\cdot \symB+\symA)$ \comm{0.35}{weak lower bound \eqref{equ:weak_lower_bound}}
\Else
\State $\symX \gets \symNumReg'/b \cdot \log(1+\symB/\symA)$ 
\comm{0.35}{strong lower bound \eqref{equ:strong_lower_bound}}
\EndIf
\algstore{myalg}
\end{algorithmic}
\end{algorithm}

\begin{algorithm}
\caption{Maximum likelihood cardinality estimation (continued).}
\begin{algorithmic}
\algrestore{myalg}
\State $\Delta\symX \gets \symX$
\State $\symStopDelta\gets\symStopEpsilon/\sqrt{\symNumReg}$
\comm{0.35}{$\symStopEpsilon = 10^{-2}$ (see \cref{sec:comp_ml_estimate})}
\While{$\Delta\symX > \symX\cdot\symStopDelta$} 
\comm{0.35}{secant method iteration}
\State $\symKappa \gets 2+\lfloor\log_2(\symX)\rfloor$
\State $\symXNormalized \gets
\symX \cdot2^{-\max(\symRegValVariate'_\text{max}, \symKappa)-1}$
\comm{0.35}{$\symXNormalized \in [0, 0.25]$}
\State $\symX''\gets \symXNormalized\cdot\symXNormalized$
\State $\symHelper \gets
\symX' - \symX''/3 + \left(\symX''\cdot \symX''\right)\cdot\left(1/45-\symX''/472.5\right)$
\comm{0.35}{Taylor approximation \eqref{equ:taylor}}
\For{$\symRegVal\gets (\symKappa - 1),\symRegValVariate'_\text{max}$}
\State $\symHelper \gets \frac{\symXNormalized+\symHelper\cdot(1-\symHelper)}{\symXNormalized+(1-\symHelper)}$
\comm{0.35}{calculate $\symHelper\!\left(\frac{\symX}{2^{\symRegVal}}\right)$, see \eqref{equ:helper_recursion1}, at this point $\symXNormalized = \frac{\symX}{2^{\symRegVal+2}}$}
\State $\symXNormalized \gets 2\symXNormalized$
\EndFor
\State $\symFuncPrime \gets \symCount\cdot\symHelper$
\comm{0.35}{compare \eqref{equ:funcprime}}
\For{$\symRegVal\gets(\symRegValVariate'_\text{max}-1),\symRegValVariate'_\text{min}$}
\State $\symHelper \gets \frac{\symXNormalized+\symHelper\cdot(1-\symHelper)}{\symXNormalized+(1-\symHelper)}$
\comm{0.35}{calculate $\symHelper\!\left(\frac{\symX}{2^{\symRegVal}}\right)$, see \eqref{equ:helper_recursion1}, at this point $\symXNormalized = \frac{\symX}{2^{\symRegVal+2}}$}
\State $\symFuncPrime\gets \symFuncPrime + \symCountVariate_\symRegVal\cdot\symHelper$
\State $\symXNormalized \gets 2\symXNormalized$
\EndFor
\State $\symFuncPrime\gets \symFuncPrime + \symX\cdot\symA$
\If{$\symFuncPrime > \symFuncPrime_\text{prev} \wedge \symNumReg' \geq\symFuncPrime$}
\State $\Delta\symX \gets \Delta\symX \cdot \frac{\symNumReg' - \symFuncPrime}{\symFuncPrime - \symFuncPrime_\text{prev}}$
\comm{0.35}{see \eqref{equ:secant_delta}}
\Else
\State $\Delta\symX \gets 0$
\EndIf
\State $\symX \gets \symX + \Delta\symX$
\State $\symFuncPrime_\text{prev}\gets \symFuncPrime$
\EndWhile
\State \Return $\symNumReg\cdot\symX$
\EndFunction
\end{algorithmic}
\end{algorithm}

>>>>

\begin{algorithm}
\caption{Compression of a $(\symPrecision, \symRegRange)$-HyperLogLog sketch with register values $\boldsymbol{\symRegValVariate}=(\symRegValVariate_1,\ldots,\symRegValVariate_\symNumReg)$ into a $(\symPrecision', \symRegRange')$-HyperLogLog sketch with $\symPrecision'\leq\symPrecision$ and $\symPrecision'+\symRegRange' \leq\symPrecision + \symRegRange$.}
\label{alg:compress}
\begin{algorithmic}
\Function {Compress}{$\boldsymbol{\symRegValVariate}$}
\comm{.4}{$\boldsymbol{\symRegValVariate}\in\lbrace 0,1,\ldots,\symRegRange+1\rbrace^\symNumReg$, $\symNumReg = 2^\symPrecision$}
\State allocate registers for new HyperLogLog sketch $\boldsymbol{\symRegValVariate}' = (\symRegValVariate_1,\ldots,\symRegValVariate_{2^{\symPrecision'}})$
\For{$\symIndexI\gets 1, 2^{\symPrecision'}$}
\State $\symB \gets (\symIndexI-1)\cdot 2^{\symPrecision-\symPrecision'}$
\State $\symIndexJ\gets 1$
\While{$\symIndexJ \leq 2^{\symPrecision-\symPrecision'}
\wedge
\symRegValVariate_{\symB+\symIndexJ}=0
$}
\State $\symIndexJ\gets\symIndexJ+1$
\EndWhile
\If{$\symIndexJ = 1$}
\State $\symRegValVariate'_\symIndexI \gets \min(\symRegValVariate_{\symB+\symIndexJ} + \symPrecision-\symPrecision',\symRegRange'+1)$
\ElsIf{$\symIndexJ\leq2^{\symPrecision-\symPrecision'}$}
\State $\langle \symBitRepA_1, \ldots, \symBitRepA_{\symPrecision-\symPrecision'}\rangle_2 \gets \symIndexJ-1$
\State $\symRegValVariate'_\symIndexI \gets 
\min(\{\symS\mid \symS \in [1, \symPrecision-\symPrecision']  \wedge  \symBitRepA_\symS = 1\} \cup \{\symRegRange'+1\})$
\Else
\State $\symRegValVariate'_\symIndexI \gets 0$
\EndIf
\EndFor
\State \Return $\boldsymbol{\symRegValVariate}'$
\comm{.4}{$\boldsymbol{\symRegValVariate}'\in\lbrace 0,1,\ldots,\symRegRange'+1\rbrace^{\symNumReg'}$, $\symNumReg' = 2^{\symPrecision'}$}
\EndFunction
\end{algorithmic}
\end{algorithm}

@@

The original estimation algorithm as presented in \cite{Flajolet2007} including corrections for small and large cardinalities is summarized by \cref{alg:estimate_original}. 
\begin{algorithm}
\caption{Original cardinality estimation algorithm for HyperLogLog sketches that use 32-bit hash values ($\symPrecision+\symRegRange = 32$) for insertion of data items \cite{Flajolet2007}.}
\label{alg:estimate_original}
\begin{algorithmic}
\Function {EstimateCardinality}{$\boldsymbol{\symRegValVariate}$}
\comm{.4}{$\boldsymbol{\symRegValVariate}\in\lbrace 0,1,\ldots,\symRegRange+1\rbrace^\symNumReg$, $\symNumReg = 2^\symPrecision$}
\State $\symCardinalityRawEstimate = \symAlpha_\symNumReg \symNumReg^2\left(\sum_{\symIndexI=1}^\symNumReg 2^{-\symRegValVariate_\symIndexI}\right)^{\!-1}$
\comm{.4}{raw estimate \eqref{equ:raw_estimator}}
\If{$\symCardinalityRawEstimate\leq \frac{5}{2}\symNumReg$}
\State $\symCountVariate_0 = \left|\{\symIndexI \vert\symRegValVariate_\symIndexI=0\}\right|$
\If{$\symCountVariate_0\neq0$}
\State \Return $\symNumReg \log(\symNumReg / \symCountVariate_0)$
\comm{.4}{small range correction \eqref{equ:linear_counting_estimator}}
\Else
\State \Return $\symCardinalityRawEstimate$
\EndIf
\ElsIf{$\symCardinalityRawEstimate\leq \frac{1}{30}2^{32}$}
\State \Return $\symCardinalityRawEstimate$
\Else
\State\Return $-2^{32}\log(1-\symCardinalityRawEstimate/2^{32})$
\comm{.4}{large range correction \eqref{equ:large_range_estimate}}
\EndIf
\EndFunction
\end{algorithmic}
\end{algorithm}
The relative estimation error for the original method is shown in \cref{fig:original_estimate}. Unfortunatley, as can be clearly seen, the ranges where the estimation error is small for $\symCardinalityRawEstimate$ and $\symCardinalityEstimate_\text{small}$ do not overlap. Therefore, the estimation error is much larger near the transition region. To reduce the estimation error for cardinalities close to this region, it was proposed to correct $\symCardinalityRawEstimate$ for bias. Empirically collected bias correction data is either stored as set of interpolation points \cite{Heule2013}, as lookup table \cite{Rhodes2015}, or as best-fitting polynomial \cite{Sanfilippo2014}. However, all these empirical approaches treat the symptom and not the cause.

@@

We can cross-check the new estimator for the linear counting case with $\symRegRange=0$. Using the identity $\symSmallCorrectionFunc(\symX) + \symLargeCorrectionFunc(\symX) = \symAlpha_\infty\symPowerSeriesFunc(\log_2(\log(1/\symX)))/\log(1/\symX)$, we get
\begin{equation}
\symPoissonRateEstimate 
= 
\frac{\symAlpha_\infty\symNumReg}{
\symSmallCorrectionFunc(\symCountVariate_0/\symNumReg)
+
\symLargeCorrectionFunc(\symCountVariate_0/\symNumReg)
}
=
\frac{\symNumReg\log\!\left(\symNumReg/\symCountVariate_0\right)}{\symPowerSeriesFunc\!\left(\log_2\!\left(\log\!\left(\symNumReg/\symCountVariate_0\right)\right)\right)}
\end{equation}
which is as expected almost identical to the linear counting estimator \eqref{equ:linear_counting_estimator}, because $\symPowerSeriesFunc(\symX)\approx 1$ (see \cref{sec:derivation_raw_estimator}).
\subsection{Improved raw estimation algorithm}
The improved raw estimator \eqref{equ:correctedestimator} leads to a new cardinality estimation algorithm for HyperLogLog sketches. \cref{alg:corrected_raw_estimation} demonstrates the numerical evaluation of the estimator. The values of functions $\symSmallCorrectionFunc$ and $\symLargeCorrectionFunc$ can be either calculated on-demand or pre-calculated. Possible values of $\symCountVariate_0$ and $\symCountVariate_{\symRegRange+1}$ are $\{0,1,\ldots, \symNumReg\}$. Therefore, if performance matters, the function values can be precalculated for all possible arguments and kept in lookup tables of size $\symNumReg+1$.
The new estimation algorithm is very elegant, because it does neither contain magic numbers nor special cases as the original algorithm.

\begin{algorithm}
\caption{Cardinality estimation based on the improved raw estimator.}
\label{alg:corrected_raw_estimation}
\begin{algorithmic}
\Function {EstimateCardinality}{$\boldsymbol{\symCountVariate}$}
\comm{.45}{$\dim \boldsymbol{\symCountVariate} = \symRegRange + 2$, $\left\Vert\boldsymbol{\symCountVariate}\right\Vert_1 = \symNumReg$}
\State $\symZ\gets\symNumReg\cdot\symLargeCorrectionFunc(1 - \symCountVariate_{\symRegRange+1}/\symNumReg)$
\comm{.45}{alternatively, take $\symNumReg\cdot\symLargeCorrectionFunc(1 - \symCountVariate_{\symRegRange+1}/\symNumReg)$ from precalculated lookup table}
\For{$\symRegVal\gets \symRegRange, 1$}
\State $\symZ\gets0.5\cdot\left(\symZ + \symCountVariate_\symRegVal\right)$
\EndFor
\State $\symZ\gets\symZ+\symNumReg\cdot\symSmallCorrectionFunc(\symCountVariate_0/\symNumReg)$
\comm{.45}{alternatively, take $\symNumReg\cdot\symSmallCorrectionFunc(\symCountVariate_0/\symNumReg)$ from precalculated lookup table}
\State \Return$\symAlpha_\infty\symNumReg^2 / \symZ$
\comm{.45}{$\symAlpha_\infty := 1/(2\log(2))$}
\EndFunction
\\
\Function {$\symSmallCorrectionFunc$}{$\symX$}
\comm{.45}{$\symX\in[0,1]$}
\If{$\symX=1$}
\State\Return$\infty$
\EndIf
\State $\symY\gets1$
\State $\symZ\gets\symX$
\Repeat
\State $\symX\gets\symX\cdot\symX$
\State $\symZ'\gets\symZ$
\State $\symZ\gets\symZ + \symX\cdot\symY$
\State $\symY\gets2\cdot\symY$
\Until{$\symZ=\symZ'$}
\State\Return$\symZ$
\EndFunction
\\
\Function {$\symLargeCorrectionFunc$}{$\symX$}
\comm{0.45}{$\symX\in[0,1]$}
\State $\symY\gets1$
\State $\symZ\gets0$
\Repeat
\State $\symX\gets\sqrt{\symX}$
\State $\symZ'\gets\symZ$
\State $\symY\gets0.5\cdot\symY$
\State $\symZ\gets\symZ + \left(1-\symX\right)\cdot\symX\cdot\symY$
\Until{$\symZ=\symZ'$}
\State\Return$\symZ$
\EndFunction
\end{algorithmic}
\end{algorithm}

@@

\begin{algorithm}
\caption{Cardinality estimation based on the maximum likelihood principle.}
\ContinuedFloat
\label{alg:estimate_ml}
\begin{algorithmic}
\Function {EstimateCardinality}{$\boldsymbol{\symCountVariate}$}
\comm{.35}{$\dim \boldsymbol{\symCountVariate} = \symRegRange + 2$, $\left\Vert\boldsymbol{\symCountVariate}\right\Vert_1 = \symNumReg$}
\State $\symRegValVariate_\text{min} \gets \min\lbrace \symRegVal\vert\symCountVariate_\symRegVal>0\rbrace$
\If{$\symRegValVariate_\text{min} > \symRegRange$}
\State\Return $\infty$
\EndIf
\State $\symRegValVariate'_\text{min} \gets \max(\symRegValVariate_\text{min}, 1)$
\State $\symRegValVariate_\text{max} \gets \max\lbrace \symRegVal\vert\symCountVariate_\symRegVal>0\rbrace$
\State $\symRegValVariate'_\text{max} \gets \min(\symRegValVariate_\text{max}, \symRegRange)$
\State $\symZ \gets 0$
\State $\symY \gets 2^{-\symRegValVariate'_\text{max}}$
\For{$\symRegVal \gets \symRegValVariate'_\text{max},\symRegValVariate'_\text{min}$}
\State $\symZ \gets \symZ + \symCountVariate_\symRegVal\cdot\symY$
\comm{0.35}{here $\symY = 2^{-\symRegVal}$}
\State $\symY \gets 2\symY$
\EndFor
\State
\comm{0.35}{here $\symZ=\sum_{\symRegVal=1}^{\symRegRange}\frac{\symCountVariate_\symRegVal}{2^\symRegVal}$}
\State $\symCount \gets \symCountVariate_{\symRegRange+1}$
\If{$\symRegRange\geq 1$}
\State $\symCount \gets \symCount + \symCountVariate_{\symRegValVariate'_\text{max}}$
\EndIf
\State $\symFuncPrime_\text{prev}\gets 0$
\State $\symA\gets \symZ + \symCountVariate_0$
\comm{0.35}{$\symA = \sum_{\symRegVal=0}^{\symRegRange}\frac{\symCountVariate_\symRegVal}{2^\symRegVal}$}
\State $\symB\gets \symZ + 
\symCountVariate_{\symRegRange+1}\cdot 2^{-\symRegRange}$
\comm{0.35}{$\symB = \sum_{\symRegVal=1}^{\symRegRange}\frac{\symCountVariate_\symRegVal}{2^\symRegVal} + \frac{\symCountVariate_{\symRegRange+1}}{2^\symRegRange}$}
\State $\symNumReg' \gets \symNumReg - \symCountVariate_0$
\If{$\symB \leq 1.5\cdot\symA$}
\State $\symX \gets \symNumReg'/(0.5\cdot \symB+\symA)$ \comm{0.35}{weak lower bound \eqref{equ:weak_lower_bound}}
\Else
\State $\symX \gets \symNumReg'/\symB \cdot \log(1+\symB/\symA)$ 
\comm{0.35}{strong lower bound \eqref{equ:strong_lower_bound}}
\EndIf
\algstore{myalg}
\end{algorithmic}
\end{algorithm}

\begin{algorithm}
\caption{Cardinality estimation based on the maximum likelihood principle (continued).}
\begin{algorithmic}
\algrestore{myalg}
\State $\Delta\symX \gets \symX$
\State $\symStopDelta\gets\symStopEpsilon/\sqrt{\symNumReg}$
\comm{0.35}{$\symStopEpsilon = 10^{-2}$ (see \cref{sec:comp_ml_estimate})}
\While{$\Delta\symX > \symX\cdot\symStopDelta$} 
\comm{0.35}{secant method iteration}
\State $\symKappa \gets 2+\lfloor\log_2(\symX)\rfloor$
\State $\symXNormalized \gets
\symX \cdot2^{-\max(\symRegValVariate'_\text{max}, \symKappa)-1}$
\comm{0.35}{$\symXNormalized \in [0, 0.25]$}
\State $\symX''\gets \symXNormalized\cdot\symXNormalized$
\State $\symHelper \gets
\symX' - \symX''/3 + \left(\symX''\cdot \symX''\right)\cdot\left(1/45-\symX''/472.5\right)$
\comm{0.35}{Taylor approximation \eqref{equ:taylor}}
\For{$\symRegVal\gets (\symKappa - 1),\symRegValVariate'_\text{max}$}
\State $\symHelper \gets \frac{\symXNormalized+\symHelper\cdot(1-\symHelper)}{\symXNormalized+(1-\symHelper)}$
\comm{0.35}{calculate $\symHelper\!\left(\frac{\symX}{2^{\symRegVal}}\right)$, see \eqref{equ:helper_recursion1}, at this point $\symXNormalized = \frac{\symX}{2^{\symRegVal+2}}$}
\State $\symXNormalized \gets 2\symXNormalized$
\EndFor
\State $\symFuncPrime \gets \symCount\cdot\symHelper$
\comm{0.35}{compare \eqref{equ:funcprime}}
\For{$\symRegVal\gets(\symRegValVariate'_\text{max}-1),\symRegValVariate'_\text{min}$}
\State $\symHelper \gets \frac{\symXNormalized+\symHelper\cdot(1-\symHelper)}{\symXNormalized+(1-\symHelper)}$
\comm{0.35}{calculate $\symHelper\!\left(\frac{\symX}{2^{\symRegVal}}\right)$, see \eqref{equ:helper_recursion1}, at this point $\symXNormalized = \frac{\symX}{2^{\symRegVal+2}}$}
\State $\symFuncPrime\gets \symFuncPrime + \symCountVariate_\symRegVal\cdot\symHelper$
\State $\symXNormalized \gets 2\symXNormalized$
\EndFor
\State $\symFuncPrime\gets \symFuncPrime + \symX\cdot\symA$
\If{$\symFuncPrime > \symFuncPrime_\text{prev} \wedge \symNumReg' \geq\symFuncPrime$}
\State $\Delta\symX \gets \Delta\symX \cdot \frac{\symNumReg' - \symFuncPrime}{\symFuncPrime - \symFuncPrime_\text{prev}}$
\comm{0.35}{see \eqref{equ:secant_delta}}
\Else
\State $\Delta\symX \gets 0$
\EndIf
\State $\symX \gets \symX + \Delta\symX$
\State $\symFuncPrime_\text{prev}\gets \symFuncPrime$
\EndWhile
\State \Return $\symNumReg\cdot\symX$
\EndFunction
\end{algorithmic}
\end{algorithm}
